**The Role of Interoception and Theory of Mind in Deception**

### 2022-08-31 paper/manuscript_v0.Rmd
editing manuscript

Much of the current research investigating the process and mechanisms involved in lying has considered capabilities such as Theory of Mind, the ability to understand and make inferences about other people's mental states and predict their behaviours [@el2017deception; @lee2021lying] . This ability is highly relevant in the context of deception, as the liar in question may utilize their inferences about the mental states of the other to predict whether the individual accepted as true the invalid information that they have presented to them [@stewart2019deception]. However, there is a growing link between other cognitive and physiological mechanisms that may play a role in both deception detection, and fabricating successful lies. This includes the interplay of mechanisms underlying increased abilities in emotional intelligence, which has been associated with increased accuracy when detecting deception [@ainley2016bodily; @stewart2019deception].

@wielgopolan2021can examined how greater awareness of one's own emotions and the emotional states of others may enhance the ability ability to accurately detect deception. Their research found a positive correlation between emotional awareness and the 'success' of a lie, suggesting that that individuals better at recognizing their own emotional state may be better at manipulating the cues that they output [@wielgopolan2021can].

Interoception is the awareness and ability to monitor the sensations and state of one's internal organs, and is closely tied to emotional and cognitive processes [@garfinkel2015knowing]. @ohira2020predictive suggests that from a predictive coding perspective, awareness of bodily signals and rapid interpretation and learning from these interoceptive signals plays a key role in the ability to regulate one's physical state.

-expand on ohira's model

This may partially explain why mindfulness practitioners tend to have greater interoceptive awareness and greater ability to regulate stress and associated physiological indicators [@ardi2021effects; @mcmahon2021mindful].

Some research has examined how higher interoceptive accuracy, measured through heart beat detection, has been associated with vasoconstriction when faced with a liar, indicating that interoception may aid in discriminating high stakes emotional lies and truths [@gunderson2021body]. It is clear that interoception may indicate on an implicit level that one is being lied to, however, there is limited research investigating how interoception may work the other way around, in favour of a liar, as @wielgopolan2021can's work appears to suggest.

The findings by @vabba2022interoceptive indicate that high cardiac interoceptive participants were more likely to be dishonest and to tell higher egoistic lies. Meanwhile, @megias2019emotionally used EEG and heart rate to find that interoceptive awareness facilitated regulation of arousal during cognitive reappraisal. There is limited research examining whether individuals with high interoceptive accuracy may be more confident in managing physiological tells associated with emotions like guilt or anxiety that the lie may be discovered through suppression or reappraisal [@fustos2013embodiment; @megias2019emotionally].

-   more theory of mind

@@

@@

Heartbeat Tracking Task (HTT)

@@

LIE Scale ????

@@

We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal during the polygraph prediction. (??)

@@

>>>>

@@

Much of the current research investigating the process and mechanisms involved in lying has considered capabilities such as Theory of Mind, the ability to understand and make inferences about other people's mental states and predict their behaviours [@el2017deception; @lee2021lying]. This ability is highly relevant in the context of deception, as the liar in question may utilize their inferences about the mental states of the other to predict whether the individual accepted as true the invalid information that they have presented to them [@stewart2019deception]. However, there is a growing link between other cognitive and physiological mechanisms that may play a role in both deception detection, and fabricating successful lies

Interoceptive awareness (IA) is the ability to monitor, interpret, integrate and regulate the sensations and state of one's internal organs, and is closely tied to emotional and cognitive processes [@garfinkel2015knowing, @chen2021emerging]. Greater IA is associated with greater self-regulation [@weiss2014interaction], especially in mindfulness practitioners who tend to have greater interoceptive awareness and greater ability to regulate stress and associated physiological indicators [@ardi2021effects; @mcmahon2021mindful].

Some research has examined how higher interoceptive accuracy, measured through heart beat detection, has been associated with vasoconstriction when faced with a liar, indicating that IA may aid in discriminating high stakes emotional lies and truths [@gunderson2021body]. It is clear that interoception may indicate on an implicit level that one is being lied to, however, there is limited research investigating how interoception may work the other way around, in favour of a liar in regulating their own physiological state that may betray their deception. The findings by @vabba2022interoceptive indicate that high cardiac interoceptive participants were more likely to be dishonest and to tell higher egoistic lies. Meanwhile, @megias2019emotionally used EEG and heart rate to find that interoceptive awareness facilitated regulation of arousal during cognitive reappraisal. There is limited research examining whether individuals with high interoceptive accuracy may be more confident in managing physiological tells associated with emotions like guilt or anxiety that the lie may be discovered through suppression or reappraisal [@fustos2013embodiment, @megias2019emotionally].

We hypothesize that participantswith greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal during the polygraph prediction. (??)

@@

Heartbeat Tracking Task (HTT) - tap one

@@

LIE Scale ???? - doms scale structure of deception

@@

@@

@durlik2014enhanced interoceptive awareness positively correlated with negative affect, anxiety, etc before public speaking etc (fear of negative eval + mood) heart beat detection more accurate during anxiety
